Appendix 2: Manuscript and Exploratory Code

### 2025-06-11 manuscript_code.Rmd
additional exploratory analysis, fixing bd02 mistake

---
title: 'Appendix 2: Manuscript and Exploratory Code'
author: "Abigail Edwards, Lauri Hietajärvi and Mikko Arvas"
date: "`r Sys.setlocale('LC_TIME', 'C'); format(Sys.time())`"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
always_allow_html: true
---

@@

```{r eval = FALSE}
data %>% filter(bd04 < 0)
data %>% filter(bd03 > 12)

@@

```{r eval = FALSE}

@@

@@

#afterdata adding additional exclusion criteria in line with preregistered exclusion for bd02 & bd03  
```

@@

#lavaan does not understand empty factor levels
data.fit <- data %>%
  #lavaan cannot deal with unordered factor with more then 2 levels
  filter(de01 == 'man' | de01 == "woman") %>% # EXCLUSION
  filter(bd01 != 'no') %>% # EXCLUSION
  mutate(
    across(isfactor, ~ fct_drop(.x)
    ) %>% 
      mutate(
        bd04 = as.numeric(bd04), # lavaan cannot handle dates
        )
  ) %>% 
  #Original paper used sum scores instead of FA. 
  mutate(
    ss_psycown = 
      across(starts_with("po"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_selfid = 
      across(starts_with("si"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_intention = 
      across(starts_with("it"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    #ss_donation as only measuring bd02
    ss_donation = 
      across(all_of(c("bd02")), ~as.numeric(.)) %>%
      rowSums(na.rm = TRUE)
    )

@@

```{r}
corrplot(cor(data.fit2,use="pairwise.complete.obs"))
```

@@

#making bd02 the only variable in donation history

po01.model <- ' po =~ po01 + po02 + po03
                si =~ si01 + si02 + si03
                it =~ it01 + it02 + it03
                ac =~ ac01 + ac02 + ac03 + ac04
                ai =~ ai01 + ai02 + ai03 + ai04 
                iv =~ iv01 + iv02 + iv03
                bd =~ bd02

@@

ind_eff := a1* d21 * b2 
'

@@

#making bd02 the only variable in donation history

po02.model <- ' 
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd02

@@

#regression
intention ~  de01 + de02 + ie01 +  cp * blooddonor + b1 * psycown + b2 * selfid
ind_eff := a1* d21 * b2
'

@@

### Exploratory Analysis (a1*b1 comparison)

@@

### exploratory analysis (multigroup analysis to explore impact of gender)

@@

#defining new model that doesn't contained de01

ea2.model <- ' 
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd02

@@

ea2.free <- sem(
  ea2.model,
  data = data.fit2,
  group = "de01",
  estimator = "MLR",
  std.lv = TRUE,
  missing = "fiml",
  fixed.x = FALSE,
  auto.fix.first = FALSE,
  group.equal = c('loadings','intercepts')
  )

summary(ea2.free)

ea2.fixed <- sem(
  ea2.model,
  data = data.fit2,
  group = "de01",
  estimator = "MLR",
  std.lv = TRUE,
  missing = "fiml",
  fixed.x = FALSE,
  auto.fix.first = FALSE,
  group.equal = c('loadings', 'intercepts', 'regressions')
)

# this is for model comparison

lavTestLRT(ea2.free, ea2.fixed)

#ea2 <- sem(
  #ea2.model,
  #data = data.fit2,
  #group = "de01",
  #estimator = "MLR",
  #std.lv = TRUE,
  #missing = "fiml",
  #fixed.x = FALSE,
  #auto.fix.first = FALSE
#)

#summary(ea2, standardized = TRUE)

>>>>

---
title: 'Appendix 2: Manuscript and Exploratory Code'
author: "Abigail Edwards, Lauri Hietajärvi and Mikko Arvas"
date: "`r Sys.setlocale('LC_TIME', 'C'); format(Sys.time())`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
always_allow_html: true
---

@@

```{r}
data %>% filter(bd04 < 0)
data %>% filter(bd03 > 12)

@@

@@

## exploratory demographics table

```{r}
library(gtsummary)
library(dplyr)
library(kableExtra)

edata_clean <- data %>%
  filter(de01 == 'man' | de01 == "woman") %>%
  rename(
    Gender = de01,
    Age = de02,
    DonationEligibility = bd01,
    DonationCount = bd02,
    DonationCountLast2years = bd03,
    WeeksFromLastDonation = bd04
  )

# Create summary table
etable1 <- tbl_summary(
  data = edata_clean,
  include = c(Age, DonationEligibility, WeeksFromLastDonation, DonationCount, DonationCountLast2years),
  by = Gender,
  statistic = list(
    all_continuous() ~ "{mean} ({sd})",   # Use mean (SD) for continuous vars
    all_categorical() ~ "{n} ({p}%)"
  )
) %>%
  add_n()  # Adds total N for each group

# Output as LaTeX
as_kable_extra(etable1, format = "latex")

@@

#afterdata adding additional exclusion criteria in line with preregistered exclusion for bd02 & bd03  
```
##unstandarsied means for reporting

data2 <- data %>%
mutate(across(c(bd02, bd03, bd04), as.numeric))

#lavaan does not understand empty factor levels
data.fit.stan <- data2 %>%
  #lavaan cannot deal with unordered factor with more then 2 levels
  filter(de01 == 'man' | de01 == "woman") %>% # EXCLUSION
  filter(bd01 != 'no') %>% # EXCLUSION
  #Original paper used sum scores instead of FA. 
  mutate(
    ss_psycown = 
      across(starts_with("po"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE),
    ss_selfid = 
      across(starts_with("si"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE),
    ss_intention = 
      across(starts_with("it"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE),
    ss_control = 
      across(starts_with("ac"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE),
    ss_knowledge = 
      across(starts_with("ai"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE),
    ss_investment = 
      across(starts_with("iv"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE)
  )

#descriptives

data.fit.stan %>%
  summarise(across(
    c(de02, ss_control, ss_knowledge, ss_investment, bd03,
      ss_psycown, ss_selfid, ss_intention),
    list(mean = ~mean(., na.rm = TRUE),
         sd = ~sd(., na.rm = TRUE))
  ))

```
```

@@

#lavaan does not understand empty factor levels
data.fit <- data %>%
  #lavaan cannot deal with unordered factor with more then 2 levels
  filter(de01 == 'man' | de01 == "woman") %>% # EXCLUSION
  filter(bd01 != 'no') %>% # EXCLUSION
  mutate(
    across(isfactor, ~ fct_drop(.x)
    ) %>% 
      mutate(
        bd04 = as.numeric(bd04), # lavaan cannot handle dates
        )
  ) %>% 
  #Original paper used sum scores instead of FA. 
  mutate(
    ss_psycown = 
      across(starts_with("po"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_selfid = 
      across(starts_with("si"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_intention = 
      across(starts_with("it"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    #ss_donation as only measuring bd03
    ss_donation = 
      across(all_of(c("bd03")), ~as.numeric(.)) %>%
      rowSums(na.rm = TRUE)
    )

@@

```{r}
corrplot(cor(data.fit2,use="pairwise.complete.obs"))

## Exploratory analysis: missingness on ei01 variable

```{r}
#, labels = c('miss', 'not'), ordered = TRUE

data.fit.miss <- data.fit2 %>%
  mutate(
    ss_control = 
      across(starts_with("ac"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_knowledge = 
      across(starts_with("ai"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    ss_investment = 
      across(starts_with("iv"),~as.numeric(.)) %>%  
      rowMeans(na.rm = TRUE) %>% 
      scale() %>% as.numeric(),
    )

data.fit.miss$ie01bin <- ifelse(is.na(data.fit2$ie01), 1, 2)
data.fit.miss$ie01bin <- factor(data.fit.miss$ie01bin, levels = c(1, 2), labels = c("miss", "not"), ordered = TRUE)
summary(data.fit.miss)

data.fit.miss <- data.fit.miss %>% 
  mutate(
    ie01bin = as.ordered(ie01bin)
  ) %>% # and neither any factors at all
mutate(
  across(everything(),as.numeric)
)  # but this also good for plotting correlations

data.fit.miss <- data.fit.miss %>%
  select(
        #ie01bin,
         #ac04,
         de01,
         de02,
         #bd02,
         #bd04,
         ss_control,
         ss_knowledge,
         ss_investment,
         bd03,
         ss_psycown,
         ss_selfid, 
         ss_intention)

library(psych)
corrplot(cor(data.fit.miss,use="pairwise.complete.obs"), method ='number')

cor_test <- corr.test(data.fit.miss, use = "pairwise", method = "pearson")

# Combine correlation and p-values into APA-style strings
apa_matrix <- matrix("", nrow=ncol(data.fit.miss), ncol=ncol(data.fit.miss))
rownames(apa_matrix) <- colnames(data.fit.miss)
colnames(apa_matrix) <- colnames(data.fit.miss)

for (i in 1:ncol(data.fit.miss)) {
  for (j in 1:ncol(data.fit.miss)) {
    r_val <- round(cor_test$r[i, j], 2)
    p_val <- cor_test$p[i, j]

    # Format p-value with APA rules
    p_str <- ifelse(p_val < .001, "< .001", paste0("= ", round(p_val, 3)))

    # Combine r and p
    apa_matrix[i, j] <- paste0(r_val, ", _p_ ", p_str)
  }
}

# View table
apa_matrix

@@

#making bd03 the only variable in donation history

po01.model <- ' po =~ po01 + po02 + po03
                si =~ si01 + si02 + si03
                it =~ it01 + it02 + it03
                ac =~ ac01 + ac02 + ac03 + ac04
                ai =~ ai01 + ai02 + ai03 + ai04 
                iv =~ iv01 + iv02 + iv03
                bd =~ bd03

@@

ind_eff := a1* d21 * b2 
ind_eff1 := a1*b1
ind_eff2 := a2*b2
'

@@

#making bd03 the only variable in donation history

po02.model <- ' 
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd03

@@

#regression
intention ~  de01 + de02 + ie01 +  cp * blooddonor + b1 * psycown + b2 * selfid
ind_eff := a1* d21 * b2
ind_eff1 := a1*b1
ind_eff2 := a2*b2
'

@@

# Exploratory Analysis 
## Factor Scores: a1*b1 comparisos

@@

# Multigroup analysis to explore impact of gender

@@

#Lauri email code

ea2.free <- '
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd03

@@

ea2.free.fit <- sem(
  ea2.free,
  data = data.fit2,
  group = "de01",
  estimator = "MLR",
  std.lv = TRUE,
  missing = "fiml",
  fixed.x = FALSE,
  auto.fix.first = FALSE,
  group.equal = c('loadings','intercepts')
  )

ea2.fixed <- '
#measurement model
psycown =~ po01 + po02 + po03
selfid =~ si01 + si02 + si03
intention =~ it01 + it02 + it03
blooddonor =~ bd03

# regressions (separate labels per group)
psycown ~ a1 * blooddonor
selfid ~ a2 * blooddonor + d21 * psycown
intention ~ cp * blooddonor + b1 * psycown + b2 * selfid

# indirect effects
ind_eff := a1 * d21 * b2
ind_eff1:= a1 * b1
ind_eff2:= a2 * b2
'

ea2.fixed.fit <- sem(
  ea2.fixed,
  data = data.fit2,
  group = "de01",
  estimator = "MLR",
  std.lv = TRUE,
  missing = "fiml",
  fixed.x = FALSE,
  auto.fix.first = FALSE,
  group.equal = c('loadings', 'intercepts')
)

# this is for model comparison
lavTestLRT(ea2.free.fit, ea2.fixed.fit)
